The Heavy Photon Search Test Detector

### 2014-04-02 hps-testrun-nim.tex
More edits based on comments from John

 Data was taken on three different converter thicknesses with a beam current varying between 
30-70~nA and repeated with the reverse field setting of the PS dipole magnet.
\begin{table}[]
\begin{center}
{\small
\begin{tabular}{|c|c|c|}
\hline
Converter thickn. & Duration &  $e^-$ on radiator \\
 (\%$X_0$) & (s) & ($\mu$C)    \\   
\hline
%\hline
1.6   & 911 &   24.4 \\ %24385.9     \\ %27 nA
%\hline
0.18   & 2640 &   193.5 \\ % 193508.9  \\  %73 nA
%\hline
0.45  & 2149 &     140.7 \\ %  140709.9  \\ %65.5 nA
%\hline
0    & 1279  &   88.1 \\ %88079.6  \\
\hline
\end{tabular}
}
\caption{Measured integrated currents for the dedicated photon runs.}
\label{tab:currents}
\end{center}
\end{table}
The photon beam line during the test run produced a relatively large fraction of pairs 
originating upstream of the converter position. This contribution was measured during data taking 
with ``empty'' converter runs i.e. removing the converter but with all other conditions the same. 
The runs taken during the time dedicated to HPS Test is summarized in Tab.~\ref{tab:currents}.

@@

By selecting \ee{} pairs from the triggered events we are able to study basic distributions of pair 
production kinematics. Pairs of opposite charge tracks, one in the top and one in the bottom half of the 
SVT, with momentum larger than 400~MeV were selected. The pair production kinematics are relatively 
well reproduced given the alignment of the tracker as shown in Fig.~\ref{fig:pair_kin}.

The expected momentum resolution from simulation is between 4-5\% 
for tracks in the momentum range of the Test run. This is significantly smaller than the accuracy at 
which we're able estimate the scale and resolution in the Test run. By looking at the agreement 
between data and simulation in the shape of the kinematic distributions for single- and two track 
events and adjusting the overall momentum scale and smearing simulated events with a 
Gaussian resolution we estimate an uncertainty of 10\% on both the momentum scale and 
resolution.
 \begin{center}
{\small
\begin{figure*}[t]
   \includegraphics[width=5cm]{figures/h_trk_top_px_h_trk_top_px_trigsel4hit_pair1351_twotrkfilt-v6-paper}
   \includegraphics[width=5cm]{figures/h_invM_h_invM_trigsel4hit_pair1351_twotrkfilt-v6-paper}
   \includegraphics[width=5cm]{figures/h_sumE_h_sumE_trigsel4hit_pair1351_twotrkfilt-v6-paper}
\caption{Kinematic distributions for e$^+$e$^-$ pairs selected by opposite charged tracks in the top and bottom half of the tracker: track momentum in the top half of the SVT (left), invariant mass (middle) and the sum of the track momentum for the pair (right).}
\label{fig:pair_kin}
\end{figure*}
}
\end{center}
In the Test run, as well as in electron running with HPS, the dominating source of uncertainty in the 
tracking and vertexing is multiple Coulomb scattering. For the vertexing performance the foremost 
difference compared to electron beam running is that the target was located approximately 67~cm 
upstream from our nominal target position; giving almost collinear tracks in the detector. The increased 
lever arm over which tracks are extrapolated widens the resolution with up to a factor of eight (depending 
on momentum) compared to what is achieved at the nominal electron target position for HPS. 
Figure~\ref{fig:impact_param} shows the horizontal and vertical positions of the extrapolated track at the 
converter position. 
{\small
	 \begin{figure}[]
 \begin{center}
	% Plot info:
	% TestRun-v6 geometry, twotrkfilter, trigsel, hps-java-1.8-SNAPSHOT, 4hits, chi2<10
	\includegraphics[width=7cm]{figures/h_trk_top_fr_conv_z_h_trk_top_conv_z_dataMC_twotrksel.png}
	\includegraphics[width=7cm]{figures/h_trk_top_fr_conv_y_h_trk_top_conv_y_dataMC_twotrksel.png}
	\caption{ Vertical (top) and horizontal (bottom) extrapolated track position at the converter position 
	taking into account the measured fringe field. }
	\label{fig:impact_param}
\end{center}
\end{figure}
}
While residual alignment show small shifts good agreement on the widths between data and simulated 
events, indicating a good understanding of our material budget and distribution in the SVT.  
%Furthermore, tails of the vertex distributions are impossible to study with the finite data sample obtained 
%in the Test run. Nevertheless, useful information can still be 
%obtained by studying the vertex distributions. Figure~\ref{fig:vtx_pos} shows the distance of closest 
%approach of the momentum vectors extrapolated in the upstream direction from our analyzing magnet, 
%taking into account the measured fringe field of the PS dipole magnet. 
 %\begin{center}
%{\small
%	 \begin{figure*}[t]
%	% Plot info:
%	% TestRun-v6 geometry, twotrkfilter, trigsel, hps-java-1.8-SNAPSHOT, 4hits, chi2<10
%	\includegraphics[width=5cm]{figures/h_vtx_fr_x_h_vtx_x_trigsel4hit_pair1351_twotrkfilt-v6-paper}
%	\includegraphics[width=5cm]{figures/h_vtx_fr_y_h_vtx_y_trigsel4hit_pair1351_twotrkfilt-v6-paper}
%	\includegraphics[width=5cm]{figures/h_vtx_fr_z_h_vtx_z_trigsel4hit_pair1351_twotrkfilt-v6-paper}
%	\caption{ Vertex position represented by the distance of closest approach of the extrapolated 
%	momentum vectors upstream of the PS dipole magnet taking into account the measured fringe field. 
%	The overall shift from zero in the x-direction is due to a 30~mrad rotation of the SVT with respect to 
%	the beam line.}
%	\label{fig:vtx_pos}
%\end{figure*}
%}
%\end{center}

@@

The ratio of the ECal cluster energy $E$ to the momentum $p$ of a matched track in the SVT was used 
to determine the conversion factors from ADC counts to energy. To compare data and simulation, all 
inoperable or noisy channels in the SVT and ECal was disabled in both data and simulation so that any 
efficiency or bias that affect the data should be reflected in the simulation. 
%Furthermore, a conversion factor calibration is performed to equalize the gains of  each channel. 
%simulation to those observed in data by comparing .  The ``weighted $E/p$'' for each crystal is %computed, 
%representing the average $E/p$ for clusters that include the crystal
% $\frac{\sum_j w_{j,i}}{\sum_j\frac{P_j}{E_j}w_{j,i}}$.
Iteratively, conversion coefficients for each crystal is adjusted until the $E$/$p$ ratio in data and 
simulation are similar. The distribution of the $E$/$p$ ratio in data and simulation are compared in 
Fig.~\ref{fig:gains}. 
{\small
\begin{figure}[]
\begin{center}
	\includegraphics[width=7cm]{figures/h_ep_data_0_h_ep_MC_0_dataMC_1351-v6-v6gains_2-trig-top-cl600reg0.pdf}
	\caption{The ECal energy over track momentum ratio ($E$/$p$) comparing data and simulation} 
	\label{fig:gains}
\end{center}
\end{figure}
}
The peak position of the distribution indicates the sampling fraction of the ECal, the fraction of the 
incident particle energy measured in the cluster. The width and tails of the distribution 
in data indicates imperfect calibration and noise of the ECal modules. This level of calibration and the 
agreement with simulation was found to be sufficient to study normalized event rates on the Test run.

@@

\subsection{Trigger Rate Comparisons}
\label{sec:mcs}
Trigger rates observed in the Test run are dominated by multiple scattered pair-produced \ee pairs in the 
converter. In simulated events, the rate of triggers depend on the modeling of the pairs angular distribution 
and the sub-sequent multiple Coulomb scattering in the converter. Rates from different converter 
thicknesses are used to study the varying multiple Coulomb scattering contribution (pair production 
angle is constant). Restricting clusters to a well calibrated region of the ECal and subtracting the 
"no converter" background we see agreement with the rates predicted by the \egs{} simulation 
program, see Tab.~\ref{tab:mcs}.
%\begin{center}
\begin{table}
{\small
\begin{tabular}{|l|c|c|c|}
\hline
\bf Converter (\%~$X_0$) & \bf 1.60 & \bf 0.45 &	\bf 0.18 \\
\hline
EGS5 &	1162 $\pm$ 112 &	255 $\pm$ 28 &	94 $\pm$ 17	\\
\hline
GEANT4 & 2633 $\pm$ 250 & 	371 $\pm$ 38 &	114 $\pm$ 18 \\
\hline
Observed 	& 1064 $\pm$ 2 & 196 $\pm$ 1 &	92 $\pm$ 1 \\						
\hline
\end{tabular}
\caption{ Observed and predicted number of events for 1~s of beam at 90~nA for three different converter 
thicknesses. The uncertainty on the prediction includes systematic uncertainties from ECal alignment, background normalization, beam current normalization and limited statistics in the simulation.}
}
\label{tab:mcs}
\end{table}
%\end{center}
This gives further confidence that the dominant source of background occupancy for HPS, multiple 
Coulomb scattered beam electrons, is well 
described~\cite{Attwood:2005zz,Shen:1978ha,Gottschalk1993467} . 
%One of the main goals of the Test run was to evaluate the description of the tails of the 
%multiple Coulomb scattering to gain further confidence in the expected detector occupancy in the full HPS 
%experiment. Those occupancies have been evaluated using \egs{}~\cite{egs5} simulation program 
%whose \moliere{} formulation of multiple scattering show good agreement with previous studies 
%for a wide range of materials and projectiles~\cite{Attwood:2005zz,Shen:1978ha,Gottschalk1993467}. 
%Despite all data taken with a photon beam in the Test run,

>>>>

 Data was taken on three different converter thicknesses with photon fluxes between 
 1.1-2.6$\times10^{8}$/s ($0.55<$E$_\gamma<5.5$~GeV) at beam currents varying between 
 30-90~nA and repeated with the reverse field setting of the PS dipole magnet.
\begin{table}[]
\begin{center}
{\small
\begin{tabular}{|c|c|c|}
\hline
Converter thickn. & Duration &  $e^-$ on radiator \\
 (\%$X_0$) & (s) & ($\mu$C)    \\   
\hline
%\hline
1.6   & 911 &   24.4 \\ %24385.9     \\ %27 nA
%\hline
0.18   & 2640 &   193.5 \\ % 193508.9  \\  %73 nA
%\hline
0.45  & 2149 &     140.7 \\ %  140709.9  \\ %65.5 nA
%\hline
0    & 1279  &   88.1 \\ %88079.6  \\
\hline
\end{tabular}
}
\caption{Measured integrated currents for the dedicated photon runs.}
\label{tab:currents}
\end{center}
\end{table}
The photon beam line during the test run produced a relatively large fraction of pairs 
originating upstream of the converter position. This contribution was measured during data taking 
with ``empty'' converter runs i.e. removing the converter but with all other conditions the same. 
The runs taken during the time dedicated to HPS Test is summarized in Tab.~\ref{tab:currents}.

@@

By selecting \ee{} pairs from the triggered events we are able to study basic distributions of pair 
production kinematics. Pairs of oppositely charged tracks, one in the top and one in the bottom half of 
the SVT, with momentum larger than 400~MeV were selected. The pair production kinematics are 
relatively well reproduced as shown in Fig.~\ref{fig:pair_kin}.

The expected momentum resolution from simulation is between 4-5\% for tracks in the momentum 
range of the Test run. By looking at the agreement 
between data and simulation in the shape of the kinematic distributions for single- and two track 
events and adjusting the overall momentum scale and smearing simulated events with a 
Gaussian resolution we estimate an agreement with the nominal scale and resolution to within 10\%.
 \begin{center}
{\small
\begin{figure*}[t]
   \includegraphics[width=5cm]{figures/h_trk_top_px_h_trk_top_px_trigsel4hit_pair1351_twotrkfilt-v6-paper}
   \includegraphics[width=5cm]{figures/h_invM_h_invM_trigsel4hit_pair1351_twotrkfilt-v6-paper}
   \includegraphics[width=5cm]{figures/h_sumE_h_sumE_trigsel4hit_pair1351_twotrkfilt-v6-paper}
\caption{Kinematic distributions for e$^+$e$^-$ pairs selected by opposite charged tracks in the top and bottom half of the tracker: track momentum in the top half of the SVT (left), invariant mass (middle) and the sum of the track momentum for the pair (right).}
\label{fig:pair_kin}
\end{figure*}
}
\end{center}
In the Test run, as well as in electron running with HPS, the dominant source of uncertainty in the 
tracking and vertexing is multiple Coulomb scattering. For the vertexing performance the foremost 
difference compared to electron beam running is that the target was located approximately 67~cm 
upstream from our nominal target position; giving almost collinear tracks in the detector. The increased 
lever arm over which tracks are extrapolated widens the resolution with up to a factor of eight (depending 
on momentum) compared to what is achieved at the nominal electron target position for HPS. 
Figure~\ref{fig:impact_param} shows the horizontal and vertical positions of the extrapolated track at the 
converter position. 
{\small
	 \begin{figure}[]
 \begin{center}
	% Plot info:
	% TestRun-v6 geometry, twotrkfilter, trigsel, hps-java-1.8-SNAPSHOT, 4hits, chi2<10
	\includegraphics[width=7cm]{figures/h_trk_top_fr_conv_z_h_trk_top_conv_z_dataMC_twotrksel.png}
	\includegraphics[width=7cm]{figures/h_trk_top_fr_conv_y_h_trk_top_conv_y_dataMC_twotrksel.png}
	\caption{ Vertical (top) and horizontal (bottom) extrapolated track position at the converter position 
	taking into account the measured fringe field. }
	\label{fig:impact_param}
\end{center}
\end{figure}
}
While residual alignment show small shifts good agreement on the widths between data and simulated 
events, indicating a good understanding of our material budget and distribution in the SVT.  Having 
the dominant contribution to the vertex resolution approximately right demonstrates that the resolution 
in HPS, with a target at 10~cm, will be as calculated.
%Furthermore, tails of the vertex distributions are impossible to study with the finite data sample obtained 
%in the Test run. Nevertheless, useful information can still be 
%obtained by studying the vertex distributions. Figure~\ref{fig:vtx_pos} shows the distance of closest 
%approach of the momentum vectors extrapolated in the upstream direction from our analyzing magnet, 
%taking into account the measured fringe field of the PS dipole magnet. 
 %\begin{center}
%{\small
%	 \begin{figure*}[t]
%	% Plot info:
%	% TestRun-v6 geometry, twotrkfilter, trigsel, hps-java-1.8-SNAPSHOT, 4hits, chi2<10
%	\includegraphics[width=5cm]{figures/h_vtx_fr_x_h_vtx_x_trigsel4hit_pair1351_twotrkfilt-v6-paper}
%	\includegraphics[width=5cm]{figures/h_vtx_fr_y_h_vtx_y_trigsel4hit_pair1351_twotrkfilt-v6-paper}
%	\includegraphics[width=5cm]{figures/h_vtx_fr_z_h_vtx_z_trigsel4hit_pair1351_twotrkfilt-v6-paper}
%	\caption{ Vertex position represented by the distance of closest approach of the extrapolated 
%	momentum vectors upstream of the PS dipole magnet taking into account the measured fringe field. 
%	The overall shift from zero in the x-direction is due to a 30~mrad rotation of the SVT with respect to 
%	the beam line.}
%	\label{fig:vtx_pos}
%\end{figure*}
%}
%\end{center}

@@

The ratio of the ECal cluster energy $E$ to the momentum $p$ of a matched track in the SVT was used 
to determine the conversion factors from ADC counts to energy. To compare data and simulation, all 
inoperable or noisy channels in the SVT and ECal was disabled in both data and simulation so that any 
efficiency or bias that affect the data should be reflected in the simulation. 
%Furthermore, a conversion factor calibration is performed to equalize the gains of  each channel. 
%simulation to those observed in data by comparing .  The ``weighted $E/p$'' for each crystal is %computed, 
%representing the average $E/p$ for clusters that include the crystal
% $\frac{\sum_j w_{j,i}}{\sum_j\frac{P_j}{E_j}w_{j,i}}$.
Iteratively, conversion coefficients for each crystal is adjusted until the $E$/$p$ ratio in data and 
simulation are similar. The distribution of the $E$/$p$ ratio in data and simulation are compared in 
Fig.~\ref{fig:gains}. 
{\small
\begin{figure}[]
\begin{center}
	\includegraphics[width=7cm]{figures/h_ep_data_0_h_ep_MC_0_dataMC_1351-v6-v6gains_2-trig-top-cl600reg0.pdf}
	\caption{The ECal energy over track momentum ratio ($E$/$p$) comparing data and simulation 
	for inclusive events in the top half of the ECal.} 
	\label{fig:gains}
\end{center}
\end{figure}
}
The peak position of the distribution indicates the sampling fraction of the ECal, the fraction of the 
incident particle energy measured in the cluster. The width and tails of the distribution 
in data indicates imperfect calibration and noise of the ECal modules. This level of calibration and the 
agreement with simulation was found to be sufficient to study normalized event rates on the Test run.

@@

\subsection{Trigger Rate Comparisons}
\label{sec:mcs}
Trigger rates observed in the Test run are dominated by multiple Coulomb scattered \ee pairs in the 
converter. In simulated events, the rate of triggers depend on the modeling of the pairs angular 
distribution and the subsequent multiple Coulomb scattering in the converter. Rates from different 
converter 
thicknesses are used to study the varying multiple Coulomb scattering contribution (pair production 
angle is constant). Restricting clusters to a well calibrated region of the ECal and subtracting the 
"no converter" background we see agreement with the rates predicted by the \egs{} simulation 
program, see Tab.~\ref{tab:mcs}.
%\begin{center}
\begin{table}
{\small
\begin{tabular}{|l|c|c|c|}
\hline
\bf Converter (\%~$X_0$) & \bf 1.60 & \bf 0.45 &	\bf 0.18 \\
\hline
EGS5 &	1162 $\pm$ 112 &	255 $\pm$ 28 &	94 $\pm$ 17	\\
\hline
GEANT4 & 2633 $\pm$ 250 & 	371 $\pm$ 38 &	114 $\pm$ 18 \\
\hline
Observed 	& 1064 $\pm$ 2 & 196 $\pm$ 1 &	92 $\pm$ 1 \\						
\hline
\end{tabular}
\caption{ Observed and predicted number of events for 1~s of beam at 90~nA for three different converter 
thicknesses. The uncertainty on the prediction includes systematic uncertainties from ECal alignment, background normalization, beam current normalization and limited statistics in the simulation.}
}
\label{tab:mcs}
\end{table}
%\end{center}
This gives further confidence that the dominant source of background occupancy for HPS, multiple 
Coulomb scattered beam electrons, is well 
described~\cite{Attwood:2005zz,Shen:1978ha,Gottschalk1993467} . 
%One of the main goals of the Test run was to evaluate the description of the tails of the 
%multiple Coulomb scattering to gain further confidence in the expected detector occupancy in the full HPS 
%experiment. Those occupancies have been evaluated using \egs{}~\cite{egs5} simulation program 
%whose \moliere{} formulation of multiple scattering show good agreement with previous studies 
%for a wide range of materials and projectiles~\cite{Attwood:2005zz,Shen:1978ha,Gottschalk1993467}. 
%Despite all data taken with a photon beam in the Test run,
